Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-04-20 draft.tex

The progenitors of several core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images \citep{Smartt2009,Eldridge2013}, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia (e.g. \citealt{Graur2014a,Kelly2014,Graur2019}, although see \citet{McCully2014} for evidence of a progenitor for a type Iax SN). Conversely, there have been searches for surviving remnants of the binary system -- main sequence stars or WDs left behind at the SN location or kicked out at high velocity \citep[e.g.][]{Schaefer2012,Ruiz-Lapuente2018,Kerzendorf2018,Kerzendorf2019}. Promising recent results include that from \citet{Shen2018} who discovered three high-velocity WDs consistent with models of double-degenerate, double-detonation SNe Ia. However, without an unambiguous observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from providing a significant contribution to the total population of SNe Ia.

@@

Given the training set of synthetic light curves, SNN is able to return a classification (SN Ia, CCSN, or peculiar) for observed light curves. Before classifying, we remove transients with variability in multiple seasons (likely active galactic nuclei or superluminous SNe). SNN is then run on every transient with a host galaxy redshift (see Section \ref{subsubsec:sn_hosts}). We define SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$ as photometrically classified SNe Ia, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through a \texttt{SALT2} light curve fitting code and are subject to quality cuts on $x_1$ and $c$ in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve has peak brightness: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability\footnote{based on light curve fit $\chi^2$ and number of degrees of freedom.} $>0.01$. The number of SNe passing each stage of these cuts is displayed in Table 2 of \citet{Vincenzi2020} and reduces the $\sim30,000$ objects to 1604 SNe. With the addition of the photometric classifier we are left with 1441 SNe with which we proceed to inspect their host galaxies. While the inclusion of stringent cuts significantly reduces the size of the sample, it greatly enhances the purity and is in line with previous rates analyses of this nature \citep{Sullivan2006} as well as being consistent with other DES-SN analyses \citep[e.g.][]{Kelsey2021}. The contamination of this sample from CCSNe according to the simulations of \citet{Vincenzi2020} is expected to be below 3\%, which is close to that of spectroscopic samples \citep{Rubin2015}.

@@

\subsubsection{Apparent magnitude limits \label{subsubsec:mag_lims}}
As we use photometric, rather than spectroscopic, redshifts for the field galaxies, they do not suffer from spectroscopic incompleteness as the SN hosts do. Instead, the inclusion of any given galaxy in the survey area in the sample is determined simply by whether it is detected above a prescribed threshold in signal-to-noise ratio, i.e. the sample is magnitude limited modulo the cuts described in Section \ref{subsec:cuts}. To determine the apparent magnitude limit in each of the optical bands we employ the method of \citet{Johnston2007}, \citet{Teodoro2010}, and \citet{Johnston2012} (hereafter Completeness I, II, III respectively). We use the test statistics $T_{\mathrm{C}}$ and $T_{\mathrm{V}}$ from those studies, which are computed based on absolute magnitude (distance modulus). Each galaxy is ranked by its absolute magnitude (distance modulus) when compared to all other galaxies in a survey within a certain slice of distance modulus (absolute magnitude). For a galaxy of apparent magnitude $m < m_{\mathrm{lim, trial}}$ observed in a survey complete to magnitude $m_{\mathrm{lim, true}}$ where  $m_{\mathrm{lim, trial}} \leq m_{\mathrm{lim, true}}$, the expectation value of the rank is 0.5. However, if a trial limiting magnitude $m_{\mathrm{lim, trial}} \geq m_{\mathrm{lim, true}}$, there will be a lack of observed faint objects, such that the expectation value of the rank drops.

@@

Secondly, we assess the shape of our SFHs and how they depend on stellar mass by comparing with SFH measurements from the Calar Alto Legacy Integral Field spectrscopy Area survey (CALIFA) as revealed by \citet{GonzalezDelgado2017}. Our models agree qualitatively on numerous features of the SFHs: high-mass galaxies form almost all their stars at high redshift; intermediate-mass galaxies display a plateau in their SFR from $\sim4$~Gyr to the present day; low-mass galaxies show an increasing rate of star formation at $z=0$. The main discrepancy with the lowest-mass bin of the CALIFA sample $8.6 <\log (M_*/\mathrm{M}_{\odot})<9.8$, which as well as an increasing SFR at low redshift shows a second peak of star formation at large lookback times, which we do not see in our model. However, this mass bin spans over one order of magnitude in stellar mass. Upon closer inspection of our model, the lower edge of this mass bin corresponds to galaxies that form at a lookback time of 5~Gyr and display a rising SFH. At the upper edge, galaxies do indeed form at times $>10$~Gyr and display an early peak followed by exponential decline. The combination of these different model SFHs in the low-mass bin thus explain the dual-peaked nature of the observed SFHs in CALIFA.

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-1.14\pm0.05$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2011,Graur2014,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}.

>>>>

The progenitors of several core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images \citep{Smartt2009,Eldridge2013}, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia (e.g. \citealt{Graur2014a,Kelly2014,Graur2019}, although see \citet{McCully2014} for evidence of a progenitor for a type Iax SN). Conversely, there have been searches for surviving remnants of the binary system -- main sequence stars or WDs left behind at the SN location or kicked out at high velocity \citep[e.g.][]{Schaefer2012,Ruiz-Lapuente2018,Kerzendorf2018,Kerzendorf2019}. Promising recent results include that from \citet{Shen2018} who discovered three high-velocity WDs consistent with models of double-degenerate, double-detonation SNe Ia. However, without an unambiguous observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence that any progenitor channel from provides a significant contribution to the total population of SNe Ia.

@@

Given the training set of synthetic light curves, SNN returns a classification (SN Ia, CCSN, or peculiar) for observed light curves. Before classifying, we remove transients with variability in multiple seasons (likely active galactic nuclei or superluminous SNe). SNN is then run on every transient with a host galaxy redshift (see Section \ref{subsubsec:sn_hosts}). We define SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$ as photometrically classified SNe Ia, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through a \texttt{SALT2} light curve fitting code and are subject to quality cuts on $x_1$ and $c$ in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe, as well as removing those in extremely dusty environments. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve has peak brightness: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability\footnote{based on light curve fit $\chi^2$ and number of degrees of freedom.} $>0.01$. The number of SNe passing each stage of these cuts is displayed in Table 2 of \citet{Vincenzi2020} and reduces the $\sim30,000$ objects to 1604 SNe. With the addition of the photometric classifier we are left with 1441 SNe with which we proceed to inspect their host galaxies. While the inclusion of stringent cuts significantly reduces the size of the sample, it greatly enhances the purity and is in line with previous rates analyses of this nature \citep{Sullivan2006} as well as being consistent with other DES-SN analyses \citep[e.g.][]{Kelsey2021}. The contamination of this sample from CCSNe according to the simulations of \citet{Vincenzi2020} is expected to be below 3\%, which is close to that of spectroscopic samples \citep{Rubin2015}.

@@

\subsubsection{Apparent magnitude limits \label{subsubsec:mag_lims}}
As we use photometric, rather than spectroscopic, redshifts for the field galaxies, they do not suffer from spectroscopic incompleteness as the SN hosts do. Instead, the inclusion of any given galaxy in the survey area in the sample is determined simply by whether it is detected above a prescribed threshold in signal-to-noise ratio, i.e. the sample is magnitude limited modulo the cuts described in Section \ref{subsec:cuts}. To determine the apparent magnitude limit for galaxies in each of the optical bands we employ the method of \citet{Johnston2007}, \citet{Teodoro2010}, and \citet{Johnston2012} (hereafter Completeness I, II, III respectively). We use the test statistics $T_{\mathrm{C}}$ and $T_{\mathrm{V}}$ from those studies, which are computed based on absolute magnitude (distance modulus). Each galaxy is ranked by its absolute magnitude (distance modulus) when compared to all other galaxies in a survey within a certain slice of distance modulus (absolute magnitude). For a galaxy of apparent magnitude $m < m_{\mathrm{lim, trial}}$ observed in a survey complete to magnitude $m_{\mathrm{lim, true}}$ where  $m_{\mathrm{lim, trial}} \leq m_{\mathrm{lim, true}}$, the expectation value of the rank is 0.5. However, if a trial limiting magnitude $m_{\mathrm{lim, trial}} \geq m_{\mathrm{lim, true}}$, there will be a lack of observed faint objects, such that the expectation value of the rank drops.

@@

Secondly, we assess the shape of our SFHs and how they depend on stellar mass by comparing with SFH measurements from the Calar Alto Legacy Integral Field spectrscopy Area survey (CALIFA) as revealed by \citet{GonzalezDelgado2017}. Our models agree qualitatively on numerous features of the SFHs (their Fig. 5): high-mass galaxies form almost all their stars at high redshift; intermediate-mass galaxies display a plateau in their SFR from $\sim4$~Gyr to the present day; low-mass galaxies show an increasing rate of star formation at $z=0$. The main discrepancy with the lowest-mass bin of the CALIFA sample $8.6 <\log (M_*/\mathrm{M}_{\odot})<9.8$, which as well as an increasing SFR at low redshift shows a second peak of star formation at large lookback times, which we do not see in our model. However, this mass bin spans over one order of magnitude in stellar mass. Upon closer inspection of our model, the lower edge of this mass bin corresponds to galaxies that form at a lookback time of 5~Gyr and display a rising SFH. At the upper edge, galaxies do indeed form at times $>10$~Gyr and display an early peak followed by exponential decline. The combination of these different model SFHs in the low-mass bin thus explain the dual-peaked nature of the observed SFHs in CALIFA.

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-1.14\pm0.05$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2011,Graur2014,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}. The result is also qualitatively consistent with that of \citet{Strolger2020a} who fit an exponential function DTD rather than a power law. However, their best fit model indicated a higher SN rate at long delay times than a $\tau^{-1}$ DTD, which is opposite to the mild preference of our fit which is steeper and thus produces a smaller fraction of SNe at long delay times.
